---
title: "COVID-19 Data Analysis and Forecasting"
author: "Mark"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Install and Load Required Packages

```{r packages}
# install.packages("plotly")
library(tidyverse)
library(dplyr)
library(lubridate)
library(forecast)
library(ggplot2)
library(scales)
library(tseries)
```

## Load the Dataset

```{r load-data}
covid_data <- read.table('https://catalog.ourworldindata.org/garden/covid/latest/compact/compact.csv', sep=',', head=TRUE)
```

## Explore the Dataset

```{r explore-data}
head(covid_data, 5)
tail(covid_data, 10)
#sample(covid_data)
dim(covid_data)
names(covid_data)
glimpse(covid_data)
```

## Clean the Data

```{r clean-data}
covid_data$date <- as.Date(covid_data$date, format = "%Y-%m-%d")
complete_covid_data <- covid_data[!is.na(covid_data$total_cases), ]
```

## Deaths in East Africa

```{r east-africa-deaths}
complete_covid_data %>% 
  filter(country %in% c('Kenya','Uganda', 'Tanzania', 'Rwanda', 'Burundi')) %>% 
  ggplot(aes(x = reorder(country, -total_deaths), y = total_deaths)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Total COVID Deaths for countries in East Africa", x = "East Africa Countries", y = "Count of deaths") +
  scale_y_continuous(labels = comma) +
  theme_minimal()
```

## COVID-19 Cases in Kenya (2020)

```{r kenya-cases-2020}
complete_covid_data %>% 
  filter(country == "Kenya", date >= "2020-01-01" & date <= "2020-12-31") %>% 
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(linewidth = 0.5, color = 'red') +
  labs(title = "Covid cases in Kenya for 2020", x = "", y = "New COVID cases")
```

## Age Distribution of Hospital Patients During COVID

```{r age-hosp-patients}
complete_covid_data %>% 
  ggplot(aes(x = median_age, y = hosp_patients)) +
  geom_point(color = 'mediumseagreen') +
  labs(title = "Age Distribution of Hospital Patients During Covid", x = 'Age', y = 'Number of patients') +
  theme_minimal()
```

## Vaccinations Across Continents

```{r vaccinations-continents}
complete_covid_data %>% 
  filter(continent %in% c("Africa", "Asia", "Europe", "North America", "South America"), !is.na(total_vaccinations_per_hundred)) %>% 
  group_by(continent) %>%
  summarise(med_vaccinations = median(total_vaccinations_per_hundred, na.rm = TRUE)) %>% 
  ggplot(aes(x = reorder(continent, med_vaccinations), y = med_vaccinations)) +
  geom_bar(stat = "identity", fill = 'orchid') +
  labs(title = "Number of vaccinations Across Different Continents", x = "Continents", y = "Total Vaccinations per hundred") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  theme_classic()
```

## Reproduction Rate in 2020

```{r reproduction-rate}
complete_covid_data %>% 
  filter(date >= "2020-01-01" & date <= "2020-12-31") %>% 
  filter(!is.na(reproduction_rate)) %>% 
  ggplot(aes(x = date, y = reproduction_rate)) +
  geom_point(color = "steelblue") +
  labs(title = "Reproduction Rate in 2020", x = '', y = 'Reproduction Rate') +
  theme_minimal()
```

## GDP per Capita by Continent

```{r gdp-continent}
complete_covid_data %>% 
  filter(continent %in% c("Africa", "Asia", "Europe", "North America", "South America")) %>%
  group_by(continent) %>% 
  summarise(med_gdp = median(gdp_per_capita, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(continent, -med_gdp), y = med_gdp)) +
  geom_bar(stat = "identity", fill = 'mediumseagreen') +
  labs(title = "Median GDP per capita per continent in 2020", x = 'Continents', y = 'GDP Per Capita') +
  scale_y_continuous(labels = comma) +
  theme_minimal()
```

## New COVID Cases vs New Vaccinations

```{r new-cases-vs-vaccinations}
complete_covid_data %>% 
  ggplot(aes(x = new_cases, y = new_vaccinations)) +
  geom_point(color = 'steelblue') +
  labs(title = "New Covid Cases vs New Vaccines Issued", x = 'New Cases of Covid', y = 'New Vaccinations Issued') +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(labels = comma) +
  theme_classic()
```

## Forecasting COVID-19 Cases for Kenya (2025)

```{r forecast-kenya}
kenya_cases <- complete_covid_data %>%
  filter(country == "Kenya") %>%
  select(date, new_cases) %>%
  drop_na()

cases_ts <- ts(kenya_cases$new_cases, frequency = 7)
model <- auto.arima(cases_ts)
forecast_data <- forecast(model, h = 365)

autoplot(forecast_data) +
  labs(title = "Forecast of COVID-19 New Cases For 2025", y = "New Cases", x = "Number of days") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major = element_line(color = "gray80"),
        panel.grid.minor = element_blank())
```

## Export Forecast to CSV

```{r export-csv}
forecast_df <- data.frame(
  date = seq(from = max(kenya_cases$date) + 1, by = "day", length.out = 365),
  predicted_cases = as.numeric(forecast_data$mean),
  lower_80 = forecast_data$lower[, 1],
  upper_80 = forecast_data$upper[, 1],
  lower_95 = forecast_data$lower[, 2],
  upper_95 = forecast_data$upper[, 2]
)

write.csv(forecast_df, "kenya_covid_forecast_2025.csv", row.names = FALSE)
